---
title: "Reproducible Research Assignment1"
---
```{r setoptions, echo=FALSE, results='hide'}
#opts_chunk$set(echo=TRUE, results='asis')
```

## Loading and preprocessing the data
Show any code that is needed to

1. Load the data (i.e. read.csv())

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
if(!file.exists('activity.csv')){
unzip('repdata_data_activity.zip')
}
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset. 

1. Make a histogram of the total number of steps taken each day
2. Calculate and report the mean and median total number of steps taken per day

```{r}
total.steps <- tapply(data$steps, data$date, sum, na.rm=TRUE)
library(ggplot2)
qplot(total.steps, binwidth = 1000, xlab = "Histogram of the total number of steps taken each day", ylab='Frequency')
```

```r
mean(total.steps, na.rm = TRUE)
```

```
## [1] 9354
```

```r
median(total.steps, na.rm = TRUE)
```

```
## [1] 10395
```


## What is the average daily activity pattern?
1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot}
library(ggplot2)
averages <- aggregate(x = list(steps = data$steps), by = list(interval = data$interval), 
                      FUN = mean, na.rm = TRUE)
ggplot(data = averages, aes(x = interval, y = steps)) + geom_line() + xlab("Time in 5-minute interval") + 
  ylab("Average number of steps taken")
```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```r
averages[which.max(averages$steps), ]
```

```
##     interval steps
## 104      835 206.2
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```r
missing <- is.na(data$steps)
# How many missing
table(missing)
```

```
## missing
## FALSE  TRUE 
## 15264  2304
```
2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```r
# Replace each missing value with the mean value of its 5-minute interval
library(Hmisc)
filled.data <- data
filled.data$steps <- with(filled.data, impute(steps, mean))
```

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```r
total.steps <- tapply(filled.data$steps, filled.data$date, FUN = sum)
```
4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
qplot(total.steps, binwidth = 1000, xlab = "Histogram of the total number of steps taken each day", ylab='Frequency')
```

```r
mean(total.steps)
```

```
## [1] 10766.19
```

```r
median(total.steps)
```

```
## [1] 10766.19
```

Mean and median values are higher than estimates from the first part of the assignment after imputing missing data. The reason is that in the original data, there are some days with `steps` values missing (`NA`) or not recorded for 
any `interval`. The total number of steps taken in such days are set to be 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day therefore ended up with higher mean and median.

## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```r
weekday.or.weekend <- function(date) {
day <- weekdays(date)
if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
return("weekday")
else if (day %in% c("Saturday", "Sunday"))
return("weekend")
else
stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```
2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r timeseries_daytype}
averages <- aggregate(steps ~ interval + day, data = filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line(color='blue', size=.7) + facet_grid(day ~ .) + 
xlab("Time in 5-minute interval") + ylab("Number of steps") + labs(title = "Average number of steps taken per 5-minute interval")
```
